\texttt{Taxonomic supertree construction with}\texttt{\emph{ Incertae sedis}}\texttt{ taxa}

### 2018-04-19 doc/incertae_sedis.tex
lost the notes

%% LyX 2.3.0 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[english]{article}
\usepackage[T1]{fontenc}
\usepackage[latin9]{luainputenc}
\usepackage{geometry}
\geometry{verbose,tmargin=3cm,bmargin=3cm,lmargin=3cm,rmargin=3cm}
\setlength{\parskip}{6pt}
\setlength{\parindent}{0pt}
\usepackage{verbatim}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage[authoryear]{natbib}

@@

Supertree methods combine a collection of input trees with different
taxon sets into a single tree on the combined taxon set. While such
methods are usually applied to leaf-labeled phylogenies that have
been estimated from DNA sequence data, we seek to include taxonomic
information also. Taxonomies can provide a comprehensive coverage
of known taxa that inferred phylogenies cannot. Additionally, taxonomies
can provide taxonomic names, which are needed to align input phylogenies
to each other, and to allow references to existing groups.

Here we describe a supertree method that takes a taxonomy as one of
its inputs. The taxonomy has a number of features not found in the
input phylogenies. While the input trees need not cover the entire
taxon set, our method requires that the taxonomy to be comprehensive.
The taxonomy also has labels for internal nodes, whereas input phylogenies
only have labels for leaf nodes. We refer to labels of taxonomy nodes
as ``taxon names'', and require that taxon names and taxonomy nodes
have a one-to-one correspondence. The taxonomy may also contain out-degree
1 nodes. These nodes correspond to monotypic taxa which contain a
single child of lower taxonomic rank.

In practice, taxonomies may also contain nodes with uncertain placement.
These nodes are often labelled ``\emph{incertae sedis}'', which
means ``uncertain seat'' in Latin. The common interpretation of
such taxa is that they may not be moved further towards the root,
but may be moved into their sibling taxa. For example, a genus with
a sibling that is a family may be annotated as``\emph{incertae familia}'',
indicating that it is unclear which family the genus should be placed
in. \emph{Incertae sedis} taxa frequently occur when a specimen is
identified up to a given taxonomic level, but no further. Extinct
taxa are often \emph{incertae sedis}.

The Open Tree of Life project seeks to build a comprehensive supertree
covering all of life. The approach is to combine information in published
phylogenies with a comprehensive taxonomy that supplies taxon names
\citep{redelings2017supertree}. In practice, the method has been
used with the Open Tree of Life taxonomy \citep{rees2017automated}.We
describe an extension to this supertree method so that we can resolve
taxonomic uncertainty by using published phylogenies to place \emph{incertae
sedis} taxa. We describe an updated semantics for conflict between
phylogeny edges and taxonomy edges and an updated semantics for assigning
taxonomy names to nodes in the synthetic tree.

Our approach to \emph{incertae sedis} taxa enables us to use published
phylogenies to update the taxonomy by extending taxonomic groups to
include \emph{incertae sedis} taxa placed within them. The ability
to update the taxonomy is important, since the placement of \emph{incertae
sedis} taxa creates artificial conflicts with a non-updated taxonomy.
Thus, the ability to correctly assess compatibility with the taxonomy
when the taxonomy contains \emph{incertae sedis }taxa allows us to
include thousands of new taxa that were previously filtered out to
avoid conflict. It also enables us to include extinct taxa, since
many of these taxa are \emph{incertae sedis}. This makes substantial
progress towards our goal of completely representation of taxa.

@@

\paragraph{Terminology}

We briefly describe the supertree algorithm proposed in \citet{redelings2017supertree}.
This supertree algorithm takes as input a ranked list $\phyloinputs=\left\{ T_{1},\ldots,T_{n}\right\} $
of rooted input phylogenies and a single rooted taxonomy tree $\taxonomy$.
Each node in the taxonomy tree is associated with a taxon name in
the set of taxon names $\mathcal{N}$, and each taxon name is corresponds
to only one node of the taxonomy tree. We use \emph{$\mathcal{L}$
}to indicate taxon names corresponding to leaf nodes of the taxonomy
tree. Correspondence between the input trees and the taxonomy is established
by annotating leaf nodes of each input tree with taxon names. For
purposes of conflict resolution, the taxonomy tree is ranked below
all of the input trees.

In this manuscript we extend this framework by adding a set $\mathcal{U}\subseteq\mathcal{N}$
of labels that have the \emph{incertae sedis }property. The extended
taxonomy then consists of the taxonomy tree $\taxonomy$, the labels
$\mathcal{N}$, and set $\mathcal{U}$ of \emph{incertae sedis }taxa.

\paragraph{Exemplification}

While the taxonomy is comprehensive and contains all taxa, input phylogenies
only have labels on their leaf nodes, and generally include a small
fraction of the total number of taxa. Furthermore, taxa in input trees
may correspond to higher taxa on the taxonomy tree, and multiple tips
may be labeled with the same taxon name. \citet{redelings2017supertree}
described a method called 'exemplification' that reduces this problem
to a simpler form where all taxon names of input trees (i) correspond
to leaf taxa and (ii) occur only once in an input tree. For the purposes
of this paper, we therefore assume that this is the case.

\paragraph{Problem decomposition}

Given the exemplified input trees and taxonomy, the supertree algorithm
first divides the supertree problem into independent pieces, solves
the problem on these pieces, and then glues the pieces back together.
This is made possible by the assumption that any taxonomy edges that
are not contested by any of the input trees can be assumed to occur
in the supertree. We may then solve separately for what happens on
each side of such an uncontested edge. When the uncontested edge is
attached to the solution below the edge and the solution above the
edge, the complete solution is then obtained.

@@

Supertrees for each subproblem are obtained by greedily constructing
a set of rooted splits $\mathcal{C}$ that are jointly consistent
and correspond to branches of the input trees. We construct an ordered
list of splits $S(T_{i})$ by walking the branches of each tree $T_{i}$
in post-order and appending the corresponding splits to the list.
These lists are combined to form the list $\Sigma=S(T_{1})+S(T_{2})+\ldots+S(T_{n})+S(\mathbb{T})$,
where '$+$' denotes concatenation. The set $C$ is initialized to
the empty set, and we then iteratively consider each split in the
list $\Sigma$ and insert it into $C$ if the resulting set remains
jointly compatible. The BUILD algorithm is used both to check compatibility,
and to construct a compatible tree from the final set $C$.

 We seek a semantics for supertrees with \emph{incertae sedis }taxa
that satisfy the following properties:
\begin{enumerate}
\item An incertae sedis node may intrude into its siblings and their descendants.
\item Two incertae sedis siblings may not be interdigitated.
\item Any placement $T\to T^{*}$ of a tree $T$ with incertae sedis taxa
is also accessible from a tree $T^{\prime}$ that is created from
$T$ by taking a taxon, attaching it to an ancestor, and marking it
\emph{incertae sedis}.
\item The semantics of an \emph{incertae sedis} node does not depend on
assigned ranks, but only on the tree.
\item Semantics is based on deriving a split for each branch of the tree.
\end{enumerate}
Here we focus on a split-based semantics instead of a semantics based
on apply of a series of edit operations to the tree. The split-based
semantics allows us to seek trees that are simultaneously consistent
with a set of split-constraints. The use of a split-based semantics
does impose some limitations. These come primarily because in the
edit-based semantics, after we place an \emph{incertae sedis} taxon,
it is no loner \emph{incertae sedis}, whereas with the split-based
semantics taxa are either always \emph{incertae sedis} or never \emph{incertae
sedis}.

@@

@@

The reduced exclude set $\mathcal{E}(n)$ for node $n$ should therefore
not contain the descendants of its siblings marked \emph{incertae
}sedis\emph{,} but should contain the descendants of its other siblings\emph{.
}We additionally stipulate that it is not possible for \emph{incertae
sedis} taxa to intrude into their \emph{incertae sedis} siblings.
(This is not the only way to construct a split-based semantics for
\emph{incertae sedis}, but it avoids certain complications, as noted
in the discussion.) We thus define $\mathcal{I}(m,n)$ to indicate
descendants of $m$ that are excluded from $n$. Thus:
\begin{align*}
\mathcal{I}(m,n) & =\begin{cases}
\emptyset & \text{if }n\text{ is \emph{incertae sedis}\,and }m\text{ is not}\\
\mathcal{I}(m) & \text{otherwise}
\end{cases}
\end{align*}

@@

\begin{align}
\mathcal{E}(n) & =\mathcal{E}(parent(n))\cup\left[\mathcal{I}(m,n)\big|m\in siblings(n)\right].\label{eq:exclude-set-formula-1}
\end{align}
This formula allows us to compute exclude sets via a pre-order recursion
on the taxonomy tree. This recursion terminates if the exclude set
for the root node is set to $\emptyset$ as a boundary condition.

@@

OK, the idea here is that we want to avoid inter-digitation. \emph{Incertae
sedis} taxa thus exclude\emph{ incertae sedis} siblings. However,
where the first semantics allows \emph{incertae sedis }taxa to intrude
into\emph{ incertae sedis }descendants of siblings, but not allow\emph{
incertae sedis} descendants of siblings to intrude into \emph{incertae
sedis} taxa, here we do the reverse.

@@

After constructing a supertree, we still need to assign taxon names
to the supertree nodes based on the taxonomy tree in the problem.
Each taxon name $n$ corresponds to a split $S(n)=S(n)_{1}|\bullet S(n)_{2}$
on the corresponding branch of the taxonomy tree. Without \emph{incertae
sedis}, such splits are always of the form $S(n)_{1}|\bullet\mathcal{L}-S(n)_{1}$,
but with \emph{incertae sedis} taxa $S_{2}(n)$ may be smaller than
$\mathcal{L}-S(n)_{1}$.

@@

However, in the \emph{incertae sedis} framework, it is possible for
one name to apply to multiple nodes. For example, in Figure \ref{fig:One-name-can}b,
the name $A$ can apply to nodes $x$ and $y$. Here the name $A$
corresponds to the split $A1\,A2\,|\bullet B\,C$, leaving out $D$
since it is \emph{incertae sedis}. The two nodes $x$ and $y$ imply
the splits $A1\,A2\,|\bullet B\,C\,D$ and $A1\,A2\,D|\bullet B\,C$
respectively, and both of these splits imply the split $A1\,A2\,|\bullet B\,C$,
so the name $A$ can apply to both $x$ and $y$\footnote{Define ``implies'', as in the propinquity paper: $A_{1}|\bullet A_{2}\implies B_{1}|\bullet B_{2}$
means $B_{1}\subseteq A_{1}$ and $B_{2}\subseteq A_{2}$.}. This cannot happen without\emph{ incertae sedis} taxa except at
monotypic nodes. When faced with a choice about where to place a name,
our solution is to find the most tip-ward node where the name can
apply and attach the name to this node. This corresponds to a conservative
choice about whether \emph{incertae sedis} taxa should be added to
an existing taxon.
\begin{figure}
\hfill{}\subfloat[Taxonomy tree]{\includegraphics[scale=0.5]{Figures/name1nodes2/tax-shaded}

@@

\subsection{Placement}

After we have attached taxon names to the synthesis tree, we would
like to interpret the position of these names in terms of placing
\emph{incertae sedis }taxa in some spot, possibly a new spot. This
would allow us to interpret the synthesis tree as saying that phylogenetic
information has (for example) placed genus A within family B, or perhaps
in a monotypic, unnamed family that contains only $A$. The simplest
approach to placement involves noting whenever a taxon $B$ is a descendant
of a taxon $A$ on the named synthesis tree but not the taxonomy tree.
For example, in Figure \ref{fig:Handling-incertae-sedis}, taxon $C$
is a descendant of taxon $A$ in the synthetic tree (Fig. \ref{fig:Handling-incertae-sedis}d),
but not the taxonomy (Fig. \ref{fig:Handling-incertae-sedis}b). Thus,
we could say that the synthetic tree places $C$ within $A$. Furthermore,
the synthetic tree additionally places $C$ within $AB$.

@@

Synthesis with \emph{incertae sedis} taxa has the potential to resolve
uncertain taxon placements using information from phylogenies. Since
phylogenies are ranked higher than taxonomy, they may place members
of incertae sedis clades as they wish even when we do not take into
account the semantics of incertae sedis taxa. However, without the
incertae sedis semantics described above, placing a taxon $A$ within
a taxon $B$ results in conflict with taxon $B$ in the taxonomy.
This has three main effects:
\begin{enumerate}
\item The edge leading to clade $B$ is lost.
\item Any taxonomy-only taxa are placed at the MRCA of the phylogenetically
informed taxa, instead of at node $B$.
\item The name for taxon $B$ is lost.
\item We do not record $A$ as being placed within $B$.
\end{enumerate}
In order to avoid a situation where input phylogenies conflict with
a large number of taxa when \emph{incertae sedis} taxa are placed
within them, we have previously filtered \emph{incertae sedis} taxa
from the taxonomy when constructing all prior synthesis trees.

When the synthesis tree conflicts with a taxonomy node, we way that
the taxon $B$ at that node is a broken taxon. Broken taxa have two
main effects, both of which are negative. First, the name $B$ of
the broken taxon is removed from the synthesis tree. Second, the conflicting
edge for taxon $B$ is not included in the synthesis tree. This means
that any children of $B$ that are taxonomy-only will not be placed
with the children of $B$ that are mentioned in input phylogenies.
Instead the taxonomy-only children of a broken taxon move towards
the root of the synthesis tree and attach at the first higher-ranked
taxon that is an ancestor of $B$ but is not broken.

@@

\subfloat[Synthesis Tree \#1]{\includegraphics[scale=0.5]{Figures/is-clade/synth1-no-is-shaded}

}\hfill{}\subfloat[Synthesis Tree \#1]{\includegraphics[scale=0.5]{Figures/is-clade/synth1-shaded}

@@

\emph{This case makes placement more complicated. Use a sequence of
ordered placements, preorder?}
\begin{figure}

\subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/is-within-is/tree1\lyxdot tre}

>>>>

%% LyX 2.3.0 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[english]{article}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage{geometry}
\geometry{verbose,tmargin=3cm,bmargin=3cm,lmargin=3cm,rmargin=3cm}
\setlength{\parskip}{6pt}
\setlength{\parindent}{0pt}
\usepackage{verbatim}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage[authoryear]{natbib}

@@

Supertree methods combine a collection of input trees with different
taxon sets into a single tree on the combined taxon set. The methods
are usually applied to leaf-labeled phylogenetic estimates. Here we
describe a supertree approach that overcomes issues associated with
using a taxonomy as one of the inputs to a supertree method.

The Open Tree of Life project \citep{HinchliffEtAl2015} seeks to
build a comprehensive supertree \citep[see][]{redelings2017supertree}
covering all of life. The approach is to combine information in published
phylogenies with a comprehensive taxonomy that supplies taxon names.
The supertree method has been used with the Open Tree of Taxonomy
\citep[OTT hereafter, see][]{rees2017automated}, but in principle
is general. The taxonomy is a crucial input for several reasons: it
covers a wider range of species, its list of names allows for alignment
of different phylogenetic estimates to each other, and it provides
names for clades. Only a small proportion of known species have been
included in a phylogenetic analysis, thus a taxonomy is crucial for
achieving comprehensive coverage of known taxa. Phylogenetic estimates
collected from the published literature often use different names
for the same species. Lists of synonyms in OTT (and other taxonomies)
allow data curators to ``align'' input phylogenies to the taxonomy
\citep[see][ for discussion of the curation process in the Open Tree of Life project]{McTavishEtAt2015}.
This alignment is important for recognizing when two different estimates
refer to the same taxon. Biologists are familiar with names for ``higher''
taxa (taxa above the species rank). Thus, naming clades in the supertree
makes the tree easier to navigate and use.

While some taxonomies may indicate hybrid taxa, the supertree approach
described here does not consider any hybridization. Thus the taxonomic
hierarchy can be converted to a tree. While not all taxonomists believe
that named taxa should correspond to clades \citep[see, for example,][]{HorandlS2010},
the priniciples of phylogenetic classification are used by such a
large majority of practicing taxonomists, that we have chosen to treat
the taxonomic tree that mirrors the hierarchy of the taxonomy as an
estimate of phylogenetic relationships. We refer to labels of taxonomy
nodes as ``taxon names'', and assume that taxon names and taxonomy
nodes have a one-to-one correspondence. Note that the taxonomy may
also contain out-degree 1 nodes. These nodes correspond to monotypic
taxa which contain a single child of lower taxonomic rank.

The supertree method described by \citet{redelings2017supertree}
is able to use a set of phylogentic estimates and taxonomy to produce
a supertree with clades names according to the taxonomy. However,
the method was not able to accommodate the fact that taxonomies frequently
contain nodes with uncertain placement. These nodes are often labelled
``\emph{incertae sedis}'', which means ``uncertain seat'' in Latin.
The common interpretation of such taxa is that they may not be moved
further towards the root, but may be moved into their sibling taxa.
For example, a genus with a sibling that is a family may be annotated
as ``\emph{incertae familia}'', indicating that it is unclear which
family the genus should be placed in. \emph{Incertae sedis} taxa frequently
occur when a specimen is identified down to a given taxonomic level,
but no further. Extinct taxa are often \emph{incertae sedis}. OTT
is constructed from several source taxonomies, and these sources include
various ways of indicating that the position of a taxon within the
hierarchy is uncertain. OTT uses a series of ``flags'' to annotate
these taxa. The supertree of by \citet{redelings2017supertree} simply
pruned these taxa from the taxonomy and input trees. Thus, the final
supertree lacked taxa which were \emph{incertae sedis}.

Here describe a supertree method so that we can resolve taxonomic
uncertainty by using published phylogenies to place \emph{incertae
sedis} taxa. Developing this method required deciding on a set operational
semantics to be used when interpreting the \emph{incertae sedis} annotation.
These semantics affect the interpretation of the phylogenetic statements
being made by the input taxonomy and the rules for applying taxonomic
names to the final supertree. In addition to discussing how to interpret
in the \emph{incertae sedis} annotation, we describe the algorithmic
changes to the supertree pipeline that were required to adequately
represent the taxonomic uncertainty. The approach described here allows
us to include thousands of new taxa in our supertree analysis that
were previously filtered out. It will also enable the Open Tree of
Life project to include extinct taxa, since many of these taxa are
\emph{incertae sedis}. This makes substantial progress towards our
goal of comprehensive inclusion of known taxa into the Open Tree of
Life summary tree.

@@

\paragraph{Previous algorithm}

The supertree algorithm of \citet{redelings2017supertree} takes as
input a ranked list $\phyloinputs=\left\{ T_{1},\ldots,T_{n}\right\} $
of leaf-labeled rooted input phylogenies and a single rooted taxonomy
tree $\taxonomy_{p}$ that is ranked below all of the input trees.
As discussed in \citet{redelings2017supertree}, the preferred tree
would display as many of the highly ranked input splits as possible.
If computational barriers were not an issue, the method would act
by first producing a ranked list of rooted splits: $\mathcal{S}=S(T_{1})+S(T_{2})+\ldots+S(T_{n})+S(\mathbb{T})$
where $S(T_{i})$ denotes a list of non-trivial rooted splits created
from a post-order traversal of tree $i$ and '$+$' denotes concatenation.
Then the supertree would be produced in a greedy manner by iterating
over the list $S$, and adding each rooted split to a growing rooted
supertree if that split is compatible with the supertree. In other
words, the set $C$ is initialized to the empty set, and we then iteratively
consider each split in the list $\mathcal{S}$ and insert it into
$C$ if the resulting set remains jointly compatible. The BUILD algorithm\citep{AhoSSU1981}
is used both to check compatibility, and to construct a compatible
tree from the final set $C$.

Because of the large number (over 2.6 million) of tips in the full
tree, the supertree method of \citet{redelings2017supertree} uses
an approximation approach that relies on decomposition of the full
problem into subproblems. The decomposition partitioning the list
of taxonomy splits $S(\taxonomy_{p})$ into two lists: $\mathcal{Z}(\taxonomy_{p}),$the
list of splits from the taxonomy tree which are compatible with every
rooted split in the phylogenetic inputs, and $\mathcal{Y}(\taxonomy_{p})$,
the list of taxonomic splits that conflict with at least one phylogenetic
input split. Conceptually the approximate algorithm works by greedily
adding all compatible splits in altered ranked list of splits: $\mathcal{S}^{\prime}=\mathcal{Z}(\taxonomy_{p})+S(T_{1})+S(T_{2})+\ldots+S(T_{n})+\mathcal{Y}(\taxonomy_{p})$.

In practice, the decomposition allows the compatibility of splits
to be performed on smaller, relabeled trees. This can be done much
more efficiently than if compatibility was checked using the BUILD
algorithm of \citet{AhoSSU1981} on the full leaf set. Thus, the supertree
algorithm first divides the supertree problem into independent pieces
by bisecting trees at the uncontested taxa which correspond to $Z$,
then solves the supertree problem on these pieces, and finally glues
the pieces back together to produce the full tree.

The taxonomy tree, $\taxonomy_{p}$, is derived from OTT. As mentioned
above, OTT not only contains a taxonomic hierarchy, but also contains
taxonomic labels and annotations (referred to as ``flags'') on the
taxa to carry extra information. The entire hiearchy encoded by OTT
is a complete taxonomic tree, $\taxonomy^{\ast}.$ Nodes in $\taxonomy^{\ast}$
have a one-to-one correspondence with a set of labels $\mathcal{N}$
that are called taxon names. We will decompose $\mathcal{N}$ into
the taxon names for leaf taxa \emph{$\mathcal{L}$ }and higher taxa
$\mathcal{H}$. We will use $\mathcal{U}$ to refer to the subset
($\mathcal{U}\subseteq\mathcal{N}$) of taxa that are annotated with
the \emph{incertae sedis }property. Other annotations identify dubious
taxa and taxonomic names that are unwanted artifacts of the process
of building OTT. Thus, the taxonomic tree which is used as an input
to the supertree algorithm is pruned version of $\taxonomy^{\ast}$.
For our purposes, the relevant information from the taxonomy consists
of the taxonomy tree $\taxonomy^{\ast}$, the labels $\mathcal{N}$,
and set $\mathcal{U}$ of \emph{incertae sedis }taxa. In previous
work, taxa in $\mathcal{U}$ were all pruned from the taxonomic hierarchy
when $\taxonomy_{p}$ is extracted from OTT. In the present work,
we consider supertree approaches that operate on a taxonomic tree
$\taxonomy,$which is produced from $\taxonomy^{\ast}$but does not
prune \emph{incertae sedis} taxa.

Throughout we assume that leaf labels on input phylogenetic trees
are only taken from the leaf set of the taxonomic tree that is being
used and that each label occurs at most once in each input tree. \citet{redelings2017supertree}
describe a method called ``exemplification'' that ensures this is
the case. Thus pruning of the \emph{incertae sedis} taxa from the
taxonomy was also accompanied by pruning those taxa from the input
phylogenetic trees.

\paragraph{Equivalence of a taxonomy tree and a list of rooted splits}

Note that, for the purposes of the supertree method, all of the information
from the taxonomic tree is contained in the set of rooted splits.
Thus, one could imagine replacing that taxonomy, $\taxonomy_{p}$,
with a set of trees $\mathcal{R}(\taxonomy_{p})$ each of which has
a single internal edge such that each edge in $\taxonomy_{p}$is included
in one tree in $\mathcal{R}(\taxonomy_{p})$. If the order of trees
is set to be (with some abuse of notation): $\mathcal{R}(\taxonomy_{p})=\mathcal{Z}(\taxonomy_{p})+\mathcal{Y}(\taxonomy_{p})$,
then using $\mathcal{R}(\taxonomy_{p})$ in place of $\taxonomy_{p}$in
the supertree algorithm would produce the same output.

We define rooted splits for each edge by noting that removing the
edge would produce two trees. The leaf label sets of these trees the
partition the full leaf set $\mathcal{L}$ into two groups. The include
set $\mathcal{I}(e)$ which does not contain the root, and the exclude
set $\mathcal{E}(e)$ which does contain the root. Such a split may
be written
\[
\mathcal{I}(e)|\bullet\mathcal{E}(e).
\]
where the $\bullet$ indicates the root. The include set is the set
of leaves that are descendants of the daughter node of the edge. The
exclude set is the set of leaves that are not descendants of that
node. If no taxa are \emph{incertae sedis}, then the exclude set for
a node is just the total tip set minus the include set for the node:
\begin{align*}
\mathcal{E}(e) & =\mathcal{L}-\mathcal{I}(e).
\end{align*}

For a node $n$ on the tip-ward side of an edge $e$, we may also
write $\mathcal{I}(n)$ for $\mathcal{I}(e)$, and $\mathcal{E}(n)$
for $\mathcal{E}(e)$. For any two rooted splits $A=A_{1}|\bullet A_{2}$
and $B=B_{1}|\bullet B_{2}$ , we say that $A$ \emph{displays }$B$
if $B_{1}\subseteq A_{1}$ and $B_{2}\subseteq A_{2}$.

Note that when a taxon rooted at node $n$ is not marked as \emph{incertae
sedis}, then the leaf labels that are descendants of $n$ will be
in the include set for $n$ and all its ancestors and will be in the
exclude set for all other nodes/edges.

\paragraph{Information content of the \emph{incertae sedis }annotation}

In order to incorporate \emph{incertae sedis }taxa into a supertree
analyses, we must arrive at an operational definition of the meaning
of the \emph{incertae sedis} label. The core meaning of the phrase
is that the annotated taxon may actually be a member of one of the
taxa that appear as siblings to it in the heirarchical representation
of the taxonomy. In the most expansive interpretation, the author
of the taxonomy is stating that the \emph{incertae sedis} taxon could
be placed anywhere inside one of the sibling taxa or their descendants.
However, frequently only a subset of the possible placement points
would be viewed as plausible. Ideally, the \emph{incertae sedis} annotation
some taxon \emph{x} would be accompanied by a specific statement about
which other taxa might include \emph{x}. Detailed readings of the
character argumentation used by taxonomists could provide this sort
of specific information in many cases, but we lack such information
on the scale of the entire OTT. Here we consider a few operational
approaches for handling the \emph{incertae sedis} label.

In terms of rooted split interpretation, an \emph{incertae sedis}
taxon differs from a normal taxon because its leaves should not automatically
be mentioned in the exclude set of all of the edges that does not
descend from. If a taxon rooted at node $n$ is marked as \emph{incertae
sedis}, then the leaf labels that are descendants of $n$ will still
be in the include set for $n$ and all its ancestors, but these labels
will not necessarily be the exclude set for all other nodes/edges.
They may simply be omitted in some rooted split representations, resulting
in partial splits.

\paragraph{Naive interpretation: ignoring the annotation}

The most straightforward approach for increasing the taxonomic completeness
of the supertree is simply to no longer prune taxa from $\taxonomy$
because they are flagged as \emph{incertae sedis}, and then run the
rest of the pipeline of \citet{redelings2017supertree} unaltered.
However this approach results in a less resolved tree because taxonomic
vagueness (in the form of the \emph{incertae sedis} annotation) is
interpreted as an unwarranted statement that the \emph{incertae sedis}
taxa are disallowed from being members of their sibling taxa.

DO WE HAVE SOME NUMBERS ON THIS (PRESUMABLY FROM JUST RUNNING THE
ONLY PROPINQUITY WITH A DIFFERENT SET OF PRUNE FLAGS)?

\paragraph{Rich taxon-specific attachment points}

As mentioned above, the incertae sedis may mean that a taxon is unplaced,
but only a few placements within the taxonomy are possible. For example,
a genus may not be characterized as incertae sedis at the level of
the subfamily. Thus, it could be placed inside any of the hypothesized
sub-families, or within its own subfamily, but it would be unexpected
for it to be placed inside one of the other genera within a family.

In principle this information is easy to incorporate. If $\mathcal{P}$is
the set of taxa that are potential ancestors for incertae sedis taxon
$x$, then $\mathcal{L}(x)$ will occur in neither the include nor
the exclude set of any taxon in $\mathcal{P}$. Unfortunately, utilizing
taxon-specific interpretations of the incertae sedis flag relies on
character argumentation (which is lacking in OTT) or rank-based information
(which is frequently untrustworthy in OTT). Thus, here we pursue an
approach of ascribing a meaning to the \emph{incertae sedis} label
which attempts to capture the core of the idea that it articulates,
but which can be applied automatically across the taxonomy without
additional information.

@@

We seek a semantics for supertrees with \emph{incertae sedis }taxa
that satisfy the following properties:
\begin{enumerate}
\item An \emph{incertae sedis} node may intrude into its siblings and their
descendants.
\item Two \emph{incertae sedis} siblings may not be interdigitated.
\item Any placement $T\to T^{*}$ of a tree $T$ with \emph{incertae sedis}
taxa is also accessible from a tree $T^{\prime}$ that is created
from $T$ by taking a taxon, attaching it to an ancestor, and marking
it \emph{incertae sedis}.%
\begin{comment}
MTH: I don't understand this
\end{comment}
\item The semantics of an \emph{incertae sedis} node does not depend on
assigned ranks, but only on the tree.
\item Semantics is based on deriving a split for each branch of the tree.
\end{enumerate}
Here we focus on a split-based semantics instead of a semantics based
on apply of a series of edit operations to the tree. The split-based
semantics allows us to seek trees that are simultaneously consistent
with a set of split-constraints. The use of a split-based semantics
does impose some limitations. These come primarily because in the
edit-based semantics, after we place an \emph{incertae sedis} taxon,
it is no loner \emph{incertae sedis}, whereas with the split-based
semantics taxa are either always \emph{incertae sedis} or never \emph{incertae
sedis}.

@@

For any two rooted splits $A=A_{1}|\bullet A_{2}$ and $B=B_{1}|\bullet B_{2}$
, we say that $A$ \emph{displays }$B$ if $B_{1}\subseteq A_{1}$
and $B_{2}\subseteq A_{2}$.

@@

The reduced exclude set $\mathcal{E}(n)$ for node $n$ should therefore
not contain the descendants of its siblings which are marked \emph{incertae
}sedis\emph{,} but should contain the descendants of its other siblings\emph{.
}We additionally stipulate that it is not possible for \emph{incertae
sedis} taxa to intrude into their \emph{incertae sedis} siblings.
(This is not the only way to construct a split-based semantics for
\emph{incertae sedis}, but it avoids certain complications, as noted
in the discussion.) We thus define $\mathcal{I}(m,n)$ to indicate
descendants of $m$ that are excluded from $n$. Thus:
\begin{align*}
\mathcal{I}(m,n) & =\begin{cases}
\emptyset & \text{if }n\text{ is \emph{incertae sedis}\,and }m\text{ is not}\\
\mathcal{I}(m) & \text{otherwise}
\end{cases}
\end{align*}

@@

\begin{align}
\mathcal{E}(n) & =\mathcal{E}(parent(n))\cup\left[\mathcal{I}(m,n)\big|m\in siblings(n)\right].\label{eq:exclude-set-formula-1}
\end{align}
This formula allows us to compute exclude sets via a pre-order traversal
on the taxonomy tree with exclude set for the root node defined to
be $\emptyset$.

@@

OK, the idea here is that we want to avoid inter-digitation. \emph{Incertae
sedis} taxa thus exclude\emph{ incertae sedis} siblings. However,
where the first semantics allows \emph{incertae sedis }taxa to intrude
into \emph{incertae sedis} descendants of siblings, but not allow\emph{
incertae sedis} descendants of siblings to intrude into \emph{incertae
sedis} taxa, here we do the reverse.

@@

After constructing a supertree, tip nodes already have names in \emph{$\mathcal{L}$}.
However, we still need to assign higher taxon names to internal supertree
nodes based on the taxonomy tree in the problem. Each taxon name $n$
corresponds to a split $S(n)=S(n)_{1}|\bullet S(n)_{2}$ on the corresponding
branch of the taxonomy tree. Without \emph{incertae sedis}, such splits
are always of the form $S(n)_{1}|\bullet\mathcal{L}-S(n)_{1}$, but
with \emph{incertae sedis} taxa $S_{2}(n)$ may be smaller than $\mathcal{L}-S(n)_{1}$.

@@

However, in the \emph{incertae sedis} framework, it is possible for
one name to apply to multiple nodes. For example, in Figure \ref{fig:One-name-can}b,
the name $A$ can apply to nodes $x$ and $y$. Here the name $A$
corresponds to the split $A1\,A2\,|\bullet B\,C$, leaving out $D$
since it is \emph{incertae sedis}. The two nodes $x$ and $y$ display
the splits $A1\,A2\,|\bullet B\,C\,D$ and $A1\,A2\,D|\bullet B\,C$
respectively, and both of these splits display the split $A1\,A2\,|\bullet B\,C$,
so the name $A$ can apply to both $x$ and $y$. This cannot happen
without\emph{ incertae sedis} taxa except at monotypic nodes. When
faced with a choice about where to place a name, our solution is to
find the most tip-ward node where the name can apply and attach the
name to this node. This corresponds to a conservative choice about
whether \emph{incertae sedis} taxa should be added to an existing
taxon.
\begin{figure}
\hfill{}\subfloat[Taxonomy tree]{\includegraphics[scale=0.5]{Figures/name1nodes2/tax-shaded}

@@

\subsection{\label{subsec:Placement}Placement}

After we have attached taxon names to the synthesis tree, we would
like to interpret the position of these names in terms of placing
\emph{incertae sedis }taxa in a revised taxonomy. This would allow
us to interpret the synthesis tree as saying that phylogenetic information
has (for example) placed genus A within family B, or perhaps outside
of all named families. The simplest approach to placement involves
noting whenever a taxon $B$ is a descendant of a taxon $A$ on the
named synthesis tree but not the taxonomy tree. For example, in Figure
\ref{fig:Handling-incertae-sedis}, taxon $C$ is a descendant of
taxon $A$ in the synthetic tree (Fig. \ref{fig:Handling-incertae-sedis}d),
but not the taxonomy (Fig. \ref{fig:Handling-incertae-sedis}b). Thus,
we could say that the synthetic tree places $C$ within $A$. Furthermore,
the synthetic tree additionally places $C$ within $AB$.

@@

Synthesis with \emph{incertae sedis} taxa has the potential to resolve
uncertain taxon placements using information from phylogenies. Since
phylogenies are ranked higher than taxonomy, they may place members
of incertae sedis clades as they wish even when we do not take into
account the semantics of incertae sedis taxa. However, without the
incertae sedis semantics described above, placing a taxon $A$ within
a taxon $B$ results in conflict with taxon $B$ in the taxonomy.
This has three main effects:
\begin{enumerate}
\item The edge leading to clade $B$ is lost.
\item Any taxonomy-only children of $B$ are placed at the MRCA of the phylogenetically
informed taxa, instead of at node $B$.
\item The name for taxon $B$ is lost.
\item We do not record $A$ as being placed within $B$.
\end{enumerate}
In order to avoid a situation where input phylogenies conflict with
a large number of taxa when \emph{incertae sedis} taxa are placed
within them, we have previously filtered \emph{incertae sedis} taxa
from the taxonomy when constructing all prior synthesis trees.

When the synthesis tree conflicts with a taxonomy node, we say that
the taxon $B$ at that node is a broken taxon. Broken taxa have two
main effects, both of which are negative. First, the name $B$ of
the broken taxon is removed from the synthesis tree. Second, the conflicting
edge for taxon $B$ is not included in the synthesis tree. This means
that any children of $B$ that are taxonomy-only will not be placed
with the children of $B$ that are mentioned in input phylogenies.
Instead the taxonomy-only children of a broken taxon move towards
the root of the synthesis tree and attach at the first higher-ranked
taxon that is an ancestor of $B$ but is not broken.

@@

\subfloat[Synthesis Tree (no \emph{incertae sedis})]{\includegraphics[scale=0.5]{Figures/is-clade/synth1-no-is-shaded}

}\hfill{}\subfloat[Synthesis Tree (\emph{incertae sedis} aware)]{\includegraphics[scale=0.5]{Figures/is-clade/synth1-shaded}

@@

\emph{}
\begin{figure}
\subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/is-within-is/tree1\lyxdot tre}
